Scaling Automated Theory Exploration

### 2015-08-07 AI4FM/article.tex
Final?

\begin{abstract}
We investigate the \textbf{theory exploration} (TE)
paradigm for computer-assisted Mathematics and identify limitations and
improvements for current approaches. Unlike the theorem-proving paradigm,
which requires user-provided conjectures, TE performs an open-ended
search for theorems satisfying given criteria. We see promise in TE for
identifying new abstractions and connections in libraries of software
and proofs, but realising this potential requires more scalable
algorithms than presently used.
\keywords{theory exploration, Haskell}
\end{abstract}
%
\section{Introduction}

Given a signature and a set of variables (collectively a \emph{theory}), then
\emph{theory exploration} (TE) describes any process
$(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$ for producing
terms of the theory which are well-formed, provable and satisfy some criterion
referred to as ``interesting''. These conditions give rise to the following
questions, which we use to characterise TE systems:
\begin{description}
\item [Q1] \label{Q1} How do we form terms?
\item [Q2] \label{Q2} How do we guarantee well-formedness?
\item [Q3] \label{Q3} How do we prove terms?
\item [Q4] \label{Q4} What is considered ``interesting''?
\end{description}

@@

As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about
Haskell code, which are defined as ``interesting'' if they cannot be simplified
using previously discovered equations. The intuition for such criteria is to
avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
when we already know $0 + x = x$. Whilst \textsc{Q4} is elegantly implemented
with a congruence closure relation (version 1) and a term rewriting system
(version 2), the term generation for \textsc{Q1} is performed by brute-force.

@@

Currently, the most powerful TE system for Haskell is \textsc{HipSpec}. This
uses the earlier \textsc{QuickSpec} system to enumerate all type-correct
combinations of the terms in the theory, group them into equivalence
classes using the \textsc{QuickCheck} counterexample finder, then conjecture
equations relating the members of these classes. Off-the-shelf automated theorem
provers (ATPs) are used to prove these conjectures \cite{rosen2012proving}. This
approach works well as a lemma generation system, making \textsc{HipSpec} a
capable inductive theorem prover as well as a theory exploration system
\cite{claessen2013automating}.

@@

We consider the use of type systems and ATPs as satisfactory solutions to
\textbf{Q2} and \textbf{Q3}, respectively. For the other aspect, we identify
the following potential improvements:

\begin{description}
\item [Q1]
  Enumerating all type-correct terms is a brute-force solution to this question.
  Scalable alternatives to brute-force algorithms are a well-studied area of
  Artificial Intelligence and Machine Learning. In particular, there is
  potential in applying heuristic algorithms like those surveyed in
  \cite{blum2011hybrid}. We could also data-mine previous results using
  generalisation methods like anti-unification \cite{bulychev2010anti}, and
  use statistical methods to prioritise which elements of the theory we
  investigate.
\item [Q4]
  Various ``interestingness'' criteria have been proposed, for example those
  surveyed in \cite{geng2006interestingness}. Augmenting or replacing the
  criteria may be useful, for example to distinguish useful relationships from
  incidental coincidences; or to prevent surprising, insightful equations from
  being discarded simply because they are reducible.
\end{description}

We are implementing a system called \textsc{ML4HS} to investigate the use of
these ideas for a scalable automated theory exploration system for Haskell. In
its current form, we are investigating the prioritisation of theory elements by
augmenting \textsc{QuickSpec} with a pre-processor. Inspired by the use of
premise selection \cite{kuhlwein2012overview} to reduce the search space in ATP,
we select sub-sets of the given theory to explore, chosen to try and keep
together those expressions which combine in interesting ways, and to separate
those which combine in uninteresting ways.

We hypothesise that similarity-based clustering of expressions, inspired by that
of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618}, is an effective method for
performing this separation. Future experiments will test this by measuring
throughput rates of \textsc{QuickSpec} with and without the \textsc{ML4HS}
pre-processor, to see how the performance changes with the size of the theory,
the elements of the theory (especially their types), and how much reduction we
performed with our pre-processor.

>>>>

\begin{abstract}
We investigate the \textbf{theory exploration} (TE)
paradigm for computer-assisted Mathematics and identify limitations and
improvements for current approaches. Unlike the theorem-proving paradigm,
which requires user-provided conjectures, TE performs an open-ended
search for theorems satisfying given criteria. We see promise in TE for
identifying new abstractions and connections in libraries of software
and proofs, but realising this potential requires more scalable
algorithms than presently used.
%\keywords{theory exploration, Haskell}
\end{abstract}
%
\section{Introduction}

Given a signature $\Sigma$ and a set of variables $V$, we call the pair
$(\Sigma, V)$ a \emph{theory} and use \emph{theory exploration} (TE) to refer
to any process $(\Sigma, V) \overset{TE}{\rightarrow} \text{Terms}(\Sigma, V)$
for producing terms of the theory which are well-formed, provable and satisfy
some criterion referred to as ``interesting''. These conditions give rise to the
following questions, which we use to characterise TE systems:
\begin{description}
\item [Q1] \label{Q1} How do we generate terms?
\item [Q2] \label{Q2} How do we guarantee well-formedness?
\item [Q3] \label{Q3} How do we prove terms?
\item [Q4] \label{Q4} What is considered ``interesting''?
\end{description}

@@

As an example, \textsc{QuickSpec} \cite{QuickSpec} discovers equations about
Haskell code, which are defined as ``interesting'' if they cannot be simplified
using previously discovered equations. The intuition for such criteria is to
avoid special cases of known theorems, such as $0 + 0 = 0$, $0 + 1 = 1$, etc.
when we already know $0 + x = x$. Whilst \textbf{Q4} is elegantly implemented
with a congruence closure relation (version 1) and a term rewriting system
(version 2), the term generation for \textbf{Q1} is performed by brute-force.

Although \textsc{QuickSpec} only \emph{tests} its equations rather than
proving them, it is still used as the exploration component of more rigorous
systems like \textsc{HipSpec} and \textsc{Hipster}.

@@

Currently, the most powerful TE system for Haskell is \textsc{HipSpec}, which
uses off-the-shelf automated theorem provers (ATPs) to verify the conjectures of
\textsc{QuickSpec}. \textsc{QuickSpec}, in turn, enumerates all type-correct combinations
of the terms in the theory up to some depth, groups them into equivalence
classes using the \textsc{QuickCheck} counterexample finder, then conjectures
equations relating the members of these classes. This approach works well as a
lemma generation system, making \textsc{HipSpec} a capable inductive theorem
prover as well as a theory exploration system \cite{claessen2013automating}.

@@

We consider \textbf{Q2} and \textbf{Q3} to be adequately solved by the existing
use of type systems and ATPs, respectively. We identify the following potential
improvements for the other questions:

\begin{description}
\item [Q1]
  Enumerating all type-correct terms is a brute-force solution to this question.
  Scalable alternatives to brute-force algorithms are a well-studied area of
  Artificial Intelligence and Machine Learning. In particular, heuristic
  search algorithms like those surveyed in \cite{blum2011hybrid} could be used.
  We could also use Machine Learning methods to identify some sub-set of a given
  theory, to prioritise over the rest.
\item [Q4]
  Various alternative ``interestingness'' criteria have been proposed, for
  example those surveyed in \cite{geng2006interestingness}. Augmenting or
  replacing the criteria may be useful, for example to distinguish useful
  relationships from incidental coincidences; or to prevent surprising,
  insightful equations from being discarded because they can be simplified.
\end{description}

We are implementing a system called \textsc{ML4HS} to investigate these ideas.
Its current form is a pre-processor for \textsc{QuickSpec} for prioritising
theory elements. Inspired by the use of premise selection
\cite{kuhlwein2012overview} to reduce the search space in ATP,
we select sub-sets of the given theory to explore, chosen to try and keep
together those expressions which combine in interesting ways, and to separate
those which combine in uninteresting ways.

We hypothesise that similarity-based clustering of expressions, inspired by that
of \textsc{ML4PG} \cite{journals/corr/abs-1212-3618} and related work in ACL2
\cite{heras2013proof}, is an effective method for performing this separation.
Future experiments will test this by comparing the throughput of
\textsc{QuickSpec} with and without the \textsc{ML4HS} pre-processor.
